\chapter{Java Template Engines on GitHub}
\label{appendix:engines}

\begin{singlespace}
\begin{itemize}
\item httl \verb!https://github.com/httl/httl!
\item Rocker \verb!https://github.com/fizzed/rocker!
\item Trimou \verb!https://github.com/trimou/trimou! \verb!http://trimou.org/!
\item jte \verb!https://github.com/casid/jte!
\item Liqp \verb!https://github.com/bkiers/Liqp!
\item Chunk \verb!https://github.com/tomj74/chunk-templates!
\item Smarty4J \verb!https://github.com/linux-china/smarty4j!
\item Rhythm \verb!https://github.com/rythmengine/rythmengine!
\item Water \verb!https://github.com/tiagobento/watertemplate-engine!
\item Wit \verb!https://github.com/febit/wit!
\item Basis \verb!https://github.com/badlogic/basis-template!
\item japid \verb!https://github.com/branaway/Japid!
\item japid42 \verb!https://github.com/branaway/japid42!
\item gt-engine \verb!https://github.com/mbknor/gt-engine!
\item gt-engine-play2 \verb!https://github.com/mbknor/gt-engine-play2!
\item Cambridge \verb!https://github.com/erdincyilmazel/Cambridge!
\item golang-like \verb!https://github.com/proninyaroslav/java-template-engine!
\item Min Velocity \verb!https://github.com/pfmiles/min-velocity!
\item Mixer2 \verb!https://github.com/nabedge/mixer2!
\item Panettone \verb!https://github.com/caelum/vraptor-panettone!
\item Carrot \verb!https://github.com/codeka/carrot!
\item PowerStat \verb!https://github.com/PowerStat/TemplateEngine!
\item Knotx \verb!https://github.com/Knotx/knotx-template-engine!
\item Inflectible \verb!https://github.com/gvlasov/inflectible!
\item te4j \verb!https://github.com/whilein/te4j!
\item static-mustache \verb!https://github.com/sviperll/static-mustache!
\item vtte \verb!https://github.com/hgschmie/vtte!
\item jade \verb!https://github.com/dhanji/jade!
\item Soulspace \verb!https://github.com/lsolbach/TemplateEngine!
\item jtt \verb!https://github.com/tokuhirom/jtt!
\item sitebuilder \verb!https://github.com/JonathanGiles/sitebuilder!
\item succinct \verb!https://github.com/tliron/succinct!
\item jelly \verb!https://github.com/pranitbose/jelly!
\item jjenkov \verb!https://github.com/jjenkov/template-engine!
\item antlr-template \verb!https://github.com/worstenemy/antlr-template!
\item datatree-templates \verb!https://github.com/berkesa/datatree-templates!
\item Simple-Template \verb!https://github.com/jalian-systems/Simple-Template!
\item Template72 \verb!https://github.com/template72/template72!
\item stencil \verb!https://github.com/impossibl/stencil!
\item coffeemaker \verb!https://github.com/mouse0w0/coffeemaker!
\item Cedilla \verb!https://github.com/ansorre/Cedilla!
\item rstl \verb!https://github.com/rstl/rstl!
\item templar \verb!https://github.com/synapticloop/templar!
\item jhaml \verb!https://github.com/iafonov/jhaml!
\item ljincheng \verb!https://github.com/ljincheng/template!
\item disc99 \verb!https://github.com/disc99/template!
\item iNamik \verb!https://github.com/iNamik/iNamik-Template-Engine!
\item Mini-templator \verb!https://github.com/bsorrentino/minitemplator!
\item Jakotem \verb!https://github.com/moznion/jakotem!
\item swiftmarker \verb!https://github.com/swiftech/swiftmarker!
\item stuartdd \verb!https://github.com/stuartdd/template!
\item zazl \verb!https://github.com/zazl/zazl!
\item josson \verb!https://github.com/octomix/josson!
\item luther94 \verb!https://github.com/luther94/templateEngine!
\item Yan \verb!https://github.com/GmailYan/TemplateEngine!
\item Holdbelief \verb!https://github.com/holdbelief/TemplateEngine!
\item buckhx \verb!https://github.com/buckhx/TemplateEngine!
\item kettlescott \verb!https://github.com/kettlescott/TemplateEngine!
\item grub \verb!https://github.com/prezi/grub!
\item twig4j \verb!https://github.com/palmenhq/twig4j-core!
\item ngoy \verb!https://github.com/krizzdewizz/ngoy!
\item templet \verb!https://github.com/toluju/templet!
\item moennig \verb!https://github.com/rnoennig/template-engine!
\item tempera \verb!https://github.com/kasonyang/tempera!
\item antiaction \verb!https://github.com/antiaction/common-template-engine!
\item templates4j \verb!https://github.com/danishdynamite/templates4j!
\item LagoVista \verb!https://github.com/LagoVistaTechLLC/java-template!
\item maddaluno \verb!https://github.com/danielemaddaluno/simple-template-engine!
\item Debuggable \verb!https://github.com/gaelyk/debuggable-template-engine!
\item Shitstorm \verb!https://github.com/hectorlf/shitstorm-template-engine!
\item Poorman \verb!https://github.com/chtz/PoormanTemplateEngine!
\item Toy \verb!https://github.com/CHZone/toy_template_engine!
\item Tiny \verb!https://github.com/ferronrsmith/tiny-template-engine-java!
\item Albirar \verb!https://github.com/albirar/albirar-template-engine!
\item Quarkus \verb!https://github.com/DavutKeskin/Quarkus-Template-Engine!
\item Tinwatchman \verb!https://github.com/tinwatchman/Simple-Java-Template-Engine!
\item Atian \verb!https://github.com/ate47/AtianTemplateEngine!
\item adrianromero \verb!https://github.com/adrianromero/templateengine!
\item mentjes \verb!https://github.com/rnentjes/Very-simple-templates!
\item round-string \verb!https://github.com/round-lang/round-string!
\item templait \verb!https://github.com/seanrmilligan/templait!
\item trodix \verb!https://github.com/trodix/teengine!
\item nate \verb!https://github.com/SeanTAllen/nate!
\item engender \verb!https://github.com/kazimsarikaya/engender!
\item ccte \verb!https://github.com/CCLooMi/ccte!
\item yyuc \verb!https://github.com/yyuc/StringTemplate!
\item hte \verb!https://github.com/HBTGmbH/hte!
\item gastirit \verb!https://github.com/gastirit/templateengine!
\item jst \verb!https://github.com/hibnico/jst!
\item Rocheteau \verb!https://github.com/JeromeRocheteau/templater!
\item quick-termplates \verb!https://github.com/rajeshputta1983/quick-templates!
\item TEFT \verb!https://github.com/marggx/TEFT!
\item stencil \verb!https://github.com/erichonorez/stencil!
\item simple-mustache \verb!https://github.com/piotrkot/simple-mustache!
\item jetg \verb!https://github.com/fbsgen/jetg!
\item yatte \verb!https://github.com/glefur/yatte!
\item simple-templates \verb!https://github.com/sbohmann/simple-templates-java!
\item Nabucco \verb!https://github.com/NABUCCO/org.nabucco.framework.template!
\item kala \verb!https://github.com/Glavo/kala-template!
\item scalpel \verb!https://github.com/bluetorch/scalpel!
\item JTemplate \verb!https://github.com/And390/JTemplate!
\item utl \verb!https://github.com/lamerexter/utl!
\item temporize \verb!https://github.com/h34tnet/temporize!
\item Webengine \verb!https://github.com/ASofterSpace/WebEngine!
\item fronton \verb!https://github.com/skozlov/fronton!
\item Rtpl \verb!https://github.com/shenfe/Rtpl!
\item raptor \verb!https://github.com/jebl01/raptor!
\item djtemplate4j \verb!https://github.com/tanob/djtemplate4j!
\item contemplate \verb!https://github.com/almondtools/comtemplate!
\item templ8 \verb!https://github.com/mpobrien/templ8!
\item jangular \verb!https://github.com/cupmanager/jangular!
\item Hotplate \verb!https://github.com/terazzo/Hotplate!
\item Trensetim \verb!https://github.com/trensetim/java-template!
\item jaml \verb!https://github.com/zhuochun/jaml!
\item jagglate \verb!https://github.com/nobuoka/Jagglate!
\item Simplate \verb!https://github.com/SuperPagh/Simplate-engine!
\item Jango \verb!https://github.com/CptPhenomeno/Jango!
\item BenFaerber \verb!https://github.com/benfaerber/WebServer!
\item libjlte \verb!https://github.com/ergor/libjlte!
\item JTTemplate \verb!https://github.com/chen3961/JTTemplate!
\item minitemplator \verb!https://github.com/chdh/minitemplator-java!
\item omnitemplate \verb!https://github.com/omnifaces/omnitemplate!
\item temmental \verb!https://github.com/jfgiraud/temmental!
\item mistigri \verb!https://github.com/jido/mistigri-java!
\item lostTemple \verb!https://github.com/Darkylin/lostTemple!
\item KDom \verb!https://github.com/Kademi/KDom!
\item Templatefeather \verb!https://github.com/aliteralmind/templatefeather!
\item Javamarkup \verb!https://github.com/kmmanoj/JavaMarkup!
\item cupjate \verb!https://github.com/marvinsiq/cupjate!
\item java-latte \verb!https://github.com/peterhalada/java-latte!
\item jBraces \verb!https://github.com/anars/jBraces!
\item Jinjava \verb!https://github.com/HubSpot/jinjava!
\end{itemize}
\end{singlespace}